---
title: "Find dprime outliers"
author: "Gerasimos Gerardos"
date: "2026-02-11"
output: html_document
---

Script Information

Title: Find d′ Outliers
Author: Gerasimos Gerardos
Project: MeMoSLAP Project (P04–Leipzig)
Associated Paper: Task load modulates network interactions between bilateral fronto-parietal and cerebellar areas during verbal working memory

Description

This script was written by Gerasimos Gerardos for the MeMoSLAP Project (P04–Leipzig) as part of the analyses conducted for the paper Task load modulates network interactions between bilateral fronto-parietal and cerebellar areas during verbal working memory.

This script processes multiple behavioral task data files, combines them into a single cleaned dataset, and computes performance metrics for each participant and task difficulty level. It calculates accuracy, hit rates, false alarm rates, and sensitivity (d′) using signal detection theory with adjusted ratios. Based on these values, the script identifies blocks with poor performance (d′ ≤ 0) as outliers and exports them to a CSV file to use when setting nuisance regressors in later steps.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Core data handling
library(tidyverse)

# Mixed models (if needed later)
library(lme4)
library(lmerTest)
library(nlme)
```
Set working directory

```{r}
work_directory <- "/data/behavior/long_task_table" # adjust to your directrory 
setwd(work_directory)

```

```{r}
files <- list.files(work_directory, pattern = "*.csv", full.names = TRUE)

# Initialize an empty list to store data frames
data_list <- list()

# Read each file and store it in the list
for (file in files) {
  tryCatch({
    # Read each CSV file with the full path
    data <- read.csv(file, header = TRUE, sep = ",", na.strings = "NA", stringsAsFactors = FALSE)
    # Store the data frame in the list
    data_list[[file]] <- data
  }, error = function(e) {
    cat("Error in file:", file, "\n", e, "\n")
  })
}
```

Files connected:

```{r}

# Identify common columns
common_cols <- Reduce(intersect, lapply(data_list, names))

# Filter each data frame to include only common columns
filtered_data_list <- lapply(data_list, function(df) df[, common_cols, drop = FALSE])

# Combine data frames using bind_rows
Data_complete <- do.call(rbind,filtered_data_list)

## create a Data Set that ommits the NA values of the variable curr_n_back:

Data_complete_clean <- Data_complete[!is.na(Data_complete$curr_n_back), ]

# Check the structure of the new data set
#str(Data_complete_clean)

```

Accuracy:

```{r}

# Calculate accuracy, false alarm rate, and miss rate by participant and difficulty level (curr_n_back)
metrics_by_condition <- Data_complete_clean %>%
  group_by(participant, curr_n_back) %>%
  summarise(
    Correct = sum(result == "Correct", na.rm = TRUE),
    Correct_rejected = sum(result == "Okay", na.rm = TRUE),
    Miss = sum(result == "Absent", na.rm = TRUE),
    False_Alarm = sum(result == "Incorrect", na.rm = TRUE),
    Total = n()
  ) %>%
  mutate(
    Accuracy = (Correct + Correct_rejected) / Total,
    False_Alarm_rate = False_Alarm / Total,
    Miss_rate = Miss / Total
  )


```

d' analysis

```{r}
# adjusted ratios (Hautus, 1995) --> recommended by Gerry: Maybe ask him again? https://www.researchgate.net/post/What-does-a-high-D-prime-score-mean
      # https://neuropsychology.github.io/psycho.R/2018/03/29/SDT.html

# Calculate hit rate (HR) and false alarm rate (FAR) for each participant and condition
hit_false_alarm_rates <- Data_complete_clean %>%
  group_by(participant, curr_n_back, block.thisN) %>%
  summarise(
    Hits = sum(result == "Correct", na.rm = TRUE),
    Misses = sum(result == "Absent", na.rm = TRUE),
    False_Alarms = sum(result == "Incorrect", na.rm = TRUE),
    Correct_Rejections = sum(result == "Okay", na.rm = TRUE),
    Total_Target = Hits + Misses,
    Total_NonTarget = False_Alarms + Correct_Rejections,
    
    HR = (sum(result == "Correct", na.rm = TRUE) + 0.5) / (sum(result == "Correct", na.rm = TRUE) + sum(result == "Absent", na.rm = TRUE) + 1), # Adjusted Hit Rate: Hautus correction
    FAR = (sum(result == "Incorrect", na.rm = TRUE) + 0.5) / (sum(result == "Incorrect", na.rm = TRUE) + sum(result == "Okay", na.rm = TRUE) + 1)  # Adjusted False Alarm Rate: Hautus correction
  )

# Compute d' for each participant and condition
d_prime_values <- hit_false_alarm_rates %>%
  mutate(d_prime = qnorm(HR) - qnorm(FAR)) %>%
  ungroup() %>%
  select(participant, curr_n_back, d_prime, block.thisN) %>%
  distinct() # Keep unique values

d_prime_values$session[grepl("1$", d_prime_values$participant)] <- "1"
d_prime_values$session[grepl("2$", d_prime_values$participant)] <- "2"

d_prime_values$participant <- sapply(d_prime_values$participant, function(x) {
  # Extract the single match of "40xx"
  regmatches(x, regexpr("40\\d{2}", x))
})

# Print the d' values
d_prime_values

#write.csv(d_prime_values, "/results/task_analysis/d_values_raw", row.names = FALSE, col.names = TRUE)

```
```{r}
# outliar blocks
# load the raw table if raw data not available
#d_prime_values <- read.csv("/results/task_analysis/d_values_raw.csv", header = TRUE, stringsAsFactors = FALSE)

out_blocks <- d_prime_values[d_prime_values$d_prime<=0 , c("participant","session","block.thisN","curr_n_back")]
colnames(out_blocks) <-c("code","session","order","nback")
out_blocks$order <- out_blocks$order + 1 # adjust because of index 0

out_blocks <- as.data.frame(out_blocks)

out_blocks

write.csv(out_blocks, "/results/task_analysis/outliar_blocks.csv", row.names=FALSE) #change to your directory


```
